Comparing the Effectiveness of the Choropleth Map with a Hexagon Tile Map for Communicating Cancer Statistics

### 2020-01-10 paper/paper.Rmd
rearrange modelling information

<!-- Specific problem for Australia -->
Australia is an extreme case of heterogenous geographic units, with large difference between the smallest and largest geographic units across various granularities.
To communicate information regarding the small inner-city geographic units of Australian cities an alternative display must be employed. This will emphasise the value of the statistic for areas with high population. Figure \@ref(fig:liver) shows the distribution of the smoothed average amount of Liver cancer diagnoses for Australian Males from 2005-2014.
These inner-city areas are not visible on the choropleth map (a). The hexagon tile map (b) shows orange hexagons for the inner-city SA2s with higher than average levels of diagnoses in the capital cities of Melbourne and Sydney. The northern Queensland, and Northern Territory SA2's with higher than average rates are still visible but use less map space on the hexagon tile map. There are still many light blue areas in both displays.

@@

The null model imposed suggests that neighbours are related. The randomness imposed was smoothed to mirror the practice taken by the Australian Cancer Atlas. This smoothing allowed neighbours to be related to each other and show distributions similar to the Liver cancer distribution shown in \@ref(fig:liver).

@@

Replicate 3 of the All Cities distribution is shown in \@ref(fig:compare). The same data set is shown in both displays, the difference is the land area of each SA3 is coloured in the choropleth and the hexagon representing the SA3 is coloured in the Hexagon Tile Map. Many maps in the Choropleth lineup (a) are coloured yellow, two are half light blue, and two are obviously blue. Only one of the maps looks red.

```{r compare, fig.cap = "Comparing the same data presented in two map displays. The choropleth map display in (a) ", fig.height = 9, fig.width = 6}
ggdraw(ylim = c(0,2), xlim = c(0,1), clip = "on") + 
  draw_plot(rasterGrob(readPNG("figures/aus_cities_4_geo.png")), 0, 1, 1, 1) +
  draw_plot(rasterGrob(readPNG("figures/aus_cities_4_hex.png")), 0, 0, 1, 1) +
  draw_plot_label(c("a", "b"), 
    c(0.05, 0.05), c(1.95, 1), size = 20, color = "white") + 
  theme(plot.background = element_rect(fill = "black"))
```

@@

The results will be analysed using a generalised linear model, with a subject random effect to account for differences in individuals. There will be two main effects: map type and trend model, which gives the fixed effects part of the model to be

@@

where $y_{ij} = 0, 1$ whether the subject detected the data plot, $\mu$ is the overall mean, $\tau_i, i=1,2$ is the map type effect, $\delta_j$ is the trend model effect.
We are allowing for an interaction between map type and trend model. As the response for whether the participant detected the data plot is binary, a logistic model is used.

@@

<!--
The detection rate for participants reporting the real data trend model is used to calculate the accuracy of the responses of participants. The accuracy is contrasted across the two displays with the following hypothesis:

```{r ttest}
# Probability of detection using map types
test_dt <- d %>% mutate(type = factor(type, levels = c("Hex.", "Choro."))) %>% 
  t.test(detect ~ type, data = ., alternative = "greater", paired = FALSE)

```

A two sample t-test was used to contrast the difference between the detection rates for the two types of displays.
The value of `r test_dt$p.value` shows that it is very unlikely the difference in means is due to chance. The null hypothesis that the detection rate for the two groups is equal can be rejected in favour of the alternative, that the means are different.
-->

Figure \@ref{detection_compare} display the average detection rates by type of plot, separately for each trend model. Evaluations on the same data set are connected by a line segment. For both the Three Cities and All Cities trend models performance was substantially better for the hexagon tile map. Viewers detected the data plot substantially more often with the hexagon tile map. One replicate for the All Cities group, had similar detection rates for both plot types, and this is discussed in a later section. Surprisingly, participants could also detect the gradual spatial trend in the NW-SE group from the hexagon tile map. We expected that the choropleth map would be superior for the type of spatial pattern, but the data suggests the hexagon tile map performs equally as well.

Table \ref{tab:desc_stats} shows the means and standard deviations for type of plot and each trend model.

<!-- The detection rates were higher when participants viewed the hexagon tile map displays. The impact of the trend model is also considered and presented in table \@ref(tab:desc_stats).
The difference was greater when viewing a three cities trend model. When the trend model affected all cities participants also detected the data display in the lineup of null plots more often. The North West to South East data plot was detected with similar rates when viewing the choropleth and  hexagon tile map displays, though for three of the replicates the participants detected the data display more often. 
-->

@@

d %>% group_by(type, trend) %>%
  summarise(m = round(mean(detect),3),
    std.dev = paste0("(", round(sd(detect),2),")")) %>% 
  gather(stat, value, m, std.dev) %>% 
  pivot_wider(names_from = "trend", values_from = "value") %>% 
  arrange(type) %>% ungroup() %>% 
  mutate(Type = types) %>% select(Type, `NW-SE`, `Three Cities`, `All Cities`) %>%
  knitr::kable(., format = "latex", align = "lccc", booktabs = TRUE, caption = "")
```

Table XXX is a summary of the mixed effects model. Blah, blah, what we learn.

@@

The relationship between time taken to submit, and the correct detection by the participant of the real data display in a lineup is considered in Figure \@ref(fig:hist_height). The bimodal distributions are similar across the type of display and the trend model hidden in the real data plot.
The hexagon map displays have a peak for the group that took up to 5 seconds to submit a response. When viewing a choropleth map display, the second peak occured at around 20 seconds. The longest a participant took to respond was 60 seconds.

@@

The choropleth map displays that included a distribution of three cities was very difficult for participants.

@@

The most common choice made by participant was clusters, this response was expected when participants viewed the Hexagon display of the all cities and three cities distributions.

@@

The choices made by participants are examined in Figure \@ref(fig:choices).
Participants were misled by the choropleth display, but not the hexagon display for all cities displays except (2). The maps with a North West to South East trend was chosen with much greater frequency in all displays. All of three cities displays, except (4), were detected in the hexagon display. All except one lineup had at least one participant select the correct map in the lineup as shown in Figure \@ref(fig:choices).

@@

!["The lineup of choropleth map displays shows a distribution that affects all capital cities. The values for the inner city areas in the capital cities result in them being coloured red. However, the colour blue dominates the display as the large rural areas are filled. Replicate (2) for all cities."](figures/aus_cities_3_geo.png)

@@

## Modeling the difference

The likelihood of detecting the data plot in the lineup can be modelled using a generalized linear mixed effects model. 
The base R [@RCore] `glm()` function implements generalised linear models. The model used includes the two main effects map type and trend model, which gives the fixed effects model to be:

### Detection Rates

```{r modeling_glm1, results="asis"}
# Modelling 
glm1 <- glm(formula = detect ~ type * trend,
  family = binomial(link = "logit"), data = d)

preds <- predict(glm1, newdata = d, se=T)
pfit <- exp(preds$fit)/(1+exp(preds$fit))
se.bands.logit = tibble(upper = (preds$fit+2*preds$se.fit), lower = (preds$fit-2*preds$se.fit))
se.bands = exp(se.bands.logit)/(1+exp(se.bands.logit))

terms <- c("Intercept", "Hex", "Three cities", "All cities", "Hex : Three cities", "Hex : All cities")

tidy(glm1) %>%
  mutate_if(is.numeric, round, 2) %>% 
  rowwise() %>% 
  mutate(estimate = case_when(
    p.value <= 0.001 ~ paste0(estimate,"$^{***}$"),
    p.value <= 0.01 ~ paste0(estimate, "$^{** }$"), 
    p.value <= 0.05 ~ paste0(estimate, "$^{*  }$"), 
    p.value <= 0.01 ~ paste0(estimate, "$^{.  }$"),
    TRUE ~ paste0(estimate))) %>% 
  ungroup() %>% 
  mutate(term = terms) %>% 
  select(Term = term, 
    Estimate = estimate, 
    Error = std.error, 
    Stat = statistic,
    p.value) %>% knitr::kable(format = "latex", escape = FALSE, align= "rcrrr", booktabs = T)
```

```{r pred_glm1}
d_glm <- d %>% select(group:certainty, detect, detect_f) %>%
  bind_cols(as_tibble(se.bands)) %>% 
  mutate(pfit, 
    predicted = ifelse(pfit > 0.5, 1, 0),
    predicted = factor(predicted, levels = c(0,1), 
      labels = c("Predicted: No", "Predicted: Yes")))

table(d_glm$predicted, d_glm$detect_f)
```

This gives the model:

$$\widehat{y_{ij}} = 0.0217 + 0.42Hex -3.25_{three} -1.24_{all} + 2.37{Hex.three} + 1.08{Hex.all} $$

```{r glm2}
glm2 <- glm(formula = detect ~ type * trend + replicate_f,
  family = binomial(link = "logit"), data = d)

emmean2 <- emmeans(glm2, c("trend", "type", "replicate_f"),
                        type = "response")

int_2 <- confint(emmean2, by = c("type", "trend", "replicate_f"), adjust = "bonferroni")
int_2 %>% 
  ggplot(aes(x= replicate_f, y = prob, group = type)) + 
  geom_point() + 
  geom_line(alpha = 0.5, lty = "dashed") + 
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL, col = type), width = 0.2) +
  scale_color_manual(values = type_colours) +
  facet_grid(trend ~ .) +
  labs(x = "Replicates", y = "Detection Rate")
```

```{r lmer1, results = "asis"}
# Mixed effects models
lmer1 <- lmer(detect ~ type*trend + (1|contributor), data = d)
lmer1_d <- augment(lmer1, d)

preds_lmer <- predict(lmer1, newdata = d)
pfit_lmer <- exp(preds_lmer)/(1+exp(preds_lmer))

stargazer::stargazer(lmer1)

d_lmer <- d %>% select(group:certainty, detect, detect_f) %>%
  bind_cols(as_tibble(se.bands)) %>% 
  mutate(pfit_lmer, 
    predicted = ifelse(pfit_lmer > 0.5, 1, 0),
    predicted = factor(predicted, levels = c(0,1), labels = c("No", "Yes")))
# Hexagon maps have better chance of correct detection
# Allowing for contributor effects to vary: 0.12 strong residual
```

For a base model of Choropleth map, using a NW-SE trend model.
The detection rate for Hexagon tile maps using a NW-SE trend model changes the log odds of the detection by 0.42.

```{r lmer2, results = "asis"}
# Mixed effects models
lmer2 <- lmer(detect ~ (type|trend), data = d)
lmer2_d <- augment(lmer2, d)

preds_lmer2 <- predict(lmer2, newdata = d)
pfit_lmer2 <- exp(preds_lmer2)/(1+exp(preds_lmer2))

stargazer::stargazer(lmer2)

d_lmer2 <- d %>% select(group:certainty, detect, detect_f) %>%
  bind_cols(as_tibble(se.bands)) %>% 
  mutate(pfit_lmer2, 
    predicted = ifelse(pfit_lmer2 > 0.5, 1, 0),
    predicted = factor(predicted, levels = c(0,1), labels = c("No", "Yes")))

```

@@

```{r lmer1_model, results = "asis"}
# Mixed effects models
lmer1 <- lmer(detect ~ type*trend + (1|contributor), data = d)
tidy(lmer1) %>% mutate_if(is.numeric, .f = round, 2) %>% stargazer::stargazer(.)

# Hexagon maps have better chance of correct detection
# Allowing for contributor effects to vary: 0.12 strong residual
```

@@

Supplementary Materials
============

@@

The demonstrative images of the `shinydashboard` web application display the demographic and consent page in Figure \@ref(fig:survey1).

@@

\begin{figure}[!t] 
\centering 
\includegraphics[width=2.5in]{figures/survey_choro.png}
\caption{An example of the choropleth map lineup shown in the survey tab of the shinydashboard app.}
\label{survey_choro} 
\end{figure} 
-->
## Overall Performance

@@

## Subject specific anomolies (0% detection)

@@

# Demographics of contributors
#dem_contribs %>% count(gender)
dem_contribs %>% ggplot() + geom_bar(aes(age))
dem_contribs %>% count(gender) %>% knitr::kable(format = "latex", booktabs = TRUE)
dem_contribs %>% count(education) %>% knitr::kable(format = "latex", booktabs = TRUE)
```

Acknowledgment {#acknowledgment}
==============

@@

Bibliography
=============

>>>>

<!-- Specific problem for Australia -->
Australia is an extreme case of heterogenous geographic units, with large difference between the smallest and largest geographic units across various granularities.
To communicate information regarding the small inner-city geographic units of Australian cities an alternative display must be employed. This will emphasise the value of the statistic for areas with high population. Figure \ref{fig:liver} shows the distribution of the smoothed average amount of Liver cancer diagnoses for Australian Males from 2005-2014.
These inner-city areas are not visible on the choropleth map (a). The hexagon tile map (b) shows orange hexagons for the inner-city SA2s with higher than average levels of diagnoses in the capital cities of Melbourne and Sydney. The northern Queensland, and Northern Territory SA2's with higher than average rates are still visible but use less map space on the hexagon tile map. There are still many light blue areas in both displays.

@@

The null model imposed suggests that neighbours are related. The randomness imposed was smoothed to mirror the practice taken by the Australian Cancer Atlas. This smoothing allowed neighbours to be related to each other and show distributions similar to the Liver cancer distribution shown in \ref{fig:liver}.

@@

Replicate 3 of the All Cities distribution is shown in \ref{fig:compare}. The same data set is shown in both displays, the difference is the land area of each SA3 is coloured in the choropleth and the hexagon representing the SA3 is coloured in the Hexagon Tile Map. Many maps in the Choropleth lineup (a) are coloured yellow, two are half light blue, and two are obviously blue. Only one of the maps looks red.

@@

The likelihood of detecting the data plot in the lineup can be modelled using a linear mixed effects model. 
The base R [@RCore] `glm()` function implements generalised linear models. The model used includes the two main effects map type and trend model, which gives the fixed effects model to be:

The results will be analysed using a linear mixed effects model, with a subject random effect to account for differences in individuals. There will be two main effects: map type and trend model, with an interaction between them, which gives the fixed effects part of the model to be:

@@

where $y_{ij} = 0, 1$ whether the subject detected the data plot, $\mu$ is the overall mean, $\tau_i, i=1,2$ is the map type effect, $\delta_j$ is the trend model effect.
We are allowing for an interaction between map type and trend model. As the response for whether the participant detected the data plot is binary, a logistic link model is used.

@@

Figure \ref{detection_compare} displays the average detection rates by type of plot, separately for each trend model. Evaluations on the same data set are connected by a line segment. For both the Three Cities and All Cities trend models performance was substantially better for the hexagon tile map. Viewers detected the data plot substantially more often with the hexagon tile map. One replicate for the All Cities group, had similar detection rates for both plot types, and this is discussed in the supplementary [Lineups] section. Surprisingly, participants could also detect the gradual spatial trend in the NW-SE group from the hexagon tile map. We expected that the choropleth map would be superior for the type of spatial pattern, but the data suggests the hexagon tile map performs slightly better, or equally as well.

Table \ref{tab:desc_stats} shows the means and standard deviations of the detection rate for each type of plot and each trend model. This also gives the standard errors, the smallest standard deviation is the Three Cities trend model shown in a Choropleth display. This group had a very small detection rate of 0.038.

@@

d %>% group_by(type, trend) %>%
  summarise(m = round(mean(detect),3),
    std.dev = paste0("(", as.character(round(sd(detect),3)),")")) %>% 
  gather(stat, value, m, std.dev) %>% 
  pivot_wider(names_from = "trend", values_from = "value") %>% 
  arrange(type) %>% ungroup() %>% 
  mutate(Type = types) %>% select(Type, `NW-SE`, `Three Cities`, `All Cities`) %>%
  knitr::kable(., format = "latex", align = "lccc", booktabs = TRUE, caption = "")
```

Table \ref{detect_glm1} is a summary of the generalised linear mixed effects model.
The estimate of the intercept represents the detection rate for the Choropleth display of the NW-SE trend.

```{r detect_glmer1, results="asis"}
# Mixed effects models
glmer1 <- glmer(detect ~ type*trend + (1|contributor), 
              family = binomial, data = d)

tidy(glmer1) %>%
  mutate_at(.vars = c("estimate", "std.error"), round, 2) %>% 
  mutate(p.value = round(p.value, 3)) %>% 
  rowwise() %>% 
  mutate(estimate = case_when(
  p.value <= 0.001 ~ paste0(estimate,"$^{***}$"),
  p.value <= 0.01 ~ paste0(estimate, "$^{** }$"),
  p.value <= 0.05 ~ paste0(estimate, "$^{*  }$"),
  p.value <= 0.01 ~ paste0(estimate, "$^{.  }$"),
  TRUE ~ paste0(estimate))) %>%
  ungroup() %>% 
  filter(!is.na(std.error)) %>% 
  select(Term = term, 
    Estimate = estimate, 
    `Std. Error` = std.error, 
    `P value` = p.value) %>% 
  knitr::kable(format = "latex", escape = FALSE, align= "rcrrr", booktabs = T)
```

The model summary in Table \ref{detect_glmer1} shows that the predicted log odds of the detection rate is 0.46 times that of the choropleth lineups, for the lineups of hexagon tile map. 
The baseline trend model used is the NW-SE trend, the detection rate is expected to decrease comparatively for choropleth displays of the Three Cities and All Cities trends. However, when the interaction is considered the detection rate is expected to increase for the Three Cities and All Cities trends shows in the Hexagon tile map display.

@@

The relationship between time taken to submit, and the correct detection by the participant of the real data display in a lineup is considered in Figure \ref{fig:hist_height}. The bimodal distributions are similar across the type of display and the trend model hidden in the real data plot.
The hexagon map displays have a peak for the group that took up to 5 seconds to submit a response. When viewing a choropleth map display, the second peak occured at around 20 seconds. The longest a participant took to respond was 60 seconds.

@@

The choropleth map displays that included a distribution of Three Cities was very difficult for participants.

@@

The most common choice made by participant was clusters, this response was expected when participants viewed the Hexagon display of the All Cities and Three Cities distributions.

@@

The choices made by participants are examined in Figure \ref{fig:choices}.
Participants were misled by the choropleth display, but not the hexagon display for All Cities displays except (2). The maps with a North West to South East trend was chosen with much greater frequency in all displays. All of Three Cities displays, except (4), were detected in the hexagon display. All except one lineup had at least one participant select the correct map in the lineup as shown in Figure \ref{fig:choices}.

@@

@@

@@

@@

\newpage
Supplementary Materials
============

@@

The demonstrative images of the `shinydashboard` web application display the demographic and consent page in Figure \ref{fig:survey1}.

@@

\begin{figure}[!t] 
\centering 
\includegraphics[width=2.5in]{figures/survey_choro.png}
\caption{An example of the choropleth map lineup shown in the survey tab of the shinydashboard app.}
\label{survey_choro} 
\end{figure} 
-->

## Lineups

Lineups were created using the Australian Statistical Areas at Level 3.

![The lineup of choropleth map displays hides a distribution that affects all capital cities.](figures/aus_cities_3_geo.png)

![The lineup of choropleth map displays shows a distribution that affects all capital cities. The values for the inner city areas in the capital cities result in them being coloured red. However, the colour blue dominates the display as the large rural areas are filled. Replicate (2) for All Cities.](figures/aus_cities_3_hex.png)

The values for the inner city areas in the capital cities result in them being coloured red. However, the colour blue dominates the display as the large rural areas are filled. There were no other choropleth maps that were coloured as blue as map (3). This meant the detection rate was quite high, participants were able to select the correct distribution without being able to see the values of the capital cities.
The same data showed in the hexagon tile map display shows varied distributions across the lineup. They are more diverse in colour than the choropleth counterparts. The same location holds the real data plot, the red city centres are now visible and surrounded by yellow, and then blue hexagons.

## Overall Performance

@@

@@

# Demographics of contributors
#dem_contribs %>% count(gender)
dem_contribs %>% ggplot() + geom_bar(aes(age))

dem_contribs %>% count(gender) %>% knitr::kable(format = "latex", booktabs = TRUE)

dem_contribs %>% count(education) %>% knitr::kable(format = "latex", booktabs = TRUE)
```

Acknowledgment
==============

@@
